Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-10 bmc_template/manuscript.tex
working on ms

\startlocaldefs
\endlocaldefs

@@

\parttitle{Background} %if any
Clustering time-series data into discrete groups can improve 
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal heterogeneity in transition probabilities, a straightforward 
but little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers.

\parttitle{Results} %if any
Simulations and Florida panthers data showed model misspecification (omitting important 
sources of variation) can lead to overfitting and over-estimating number of behavioural states. Models incorporating temporal heterogeneity have lower number of states with slightly higher variation in short movement states, and able to make out of sample predictions that captures observed diurnal and autocorrelation patterns exhibited by Florida panthers.

\parttitle{Conclusion} %if any
Incorporating temporal heterogeneity reduce the selected number of behavioural states closer 
to a biologically interpretable level, improved goodness of fit and predictability. Our 
suggest that incorporating previously neglected structure in statistical models can
allow more accurate assessment of appropriate model complexity.

@@

Given a sequence of animal movements, movement models aim to find a parsimonious description
that can be used to understand past movements and predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \cite{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\cite{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} % 
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \cite{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour 
at particular times is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated walk model
\cite{okubo_diffusion_1980,turchin1998quantitative}.

@@

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
% BMB: pick one (perhaps include a parenthetical statement/definition
% saying that we treat these as a synonyms.  Just stringing them together
% with slashes is awkward)
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \cite{patterson_classifying_2009}, or by more
direct comparisons between inferred states and environmental
conditions \cite{fryxell_multiple_2008}. Schliehe-Diecks et al. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales \cite{gurarie2009novel}.

@@

van de Kerk et al. \cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
van de Kerk et al.'s results on optimal
model complexity might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  For large data sets, 
information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data leads to
selection of models with smaller numbers of latent states; we also
fit models to simulated data with varying numbers of latent states
and degrees of temporal heterogeneity to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

\newcommand{\obs}{\mathbf{y}}
\newcommand{\cov}{\mathbf{x}}
\newcommand{\state}{z}

@@

\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ where the covariate is $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on covariates $\cov_{1}$ and state $\state_{1}$.

\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ with covariate $\cov_{k}$ at time $t=k$.

\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ conditioned on covariates $\cov_{k}$ at state $\state_{k}$ at time $t=k$.  
\end{description}

@@

For any $n$-state HMM, we need to define a $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy. It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.

@@

where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).

\item[Quadratic transition model] 
We assume the elements of the linear predictor
are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there 
is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity 
constraint would collapse the model to a constant.

$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
$$

\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.

$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$

\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day. This model is included for comparative purposes due to the large number of parameters in the model which makes it not really practical. We only fitted up to four states using the hourly model.  
\end{description}

Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.

@@

\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall 
goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
observed data and for data simulated from the models shows how well a particular class of 
models can capture the diurnal variation in behaviour. (3) Comparing temporal autocorrelations
for the observed data and for data simulated from the models shows how well a particular class
of models captures serial correlation at both short and long scales.

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity. The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:

\begin{equation}
\textrm{Number of Free Parameter}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
\label{eq:numparam}
\end{equation}

As the number of states increases, the number of free parameters in time-homogeneous FMMs and HMMs and FMMs with temporal heterogeneity will increase linearly, whereas HMMs with temporal heterogeneity will increase quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict hourly step length and ACF 
because, while the computation is reasonably straightforward for FMMs, and manageable for homogeneous HMMs, the interaction between the geometric dwell time within each state and the temporally varying interaction probabilities makes it unreasonably complex. 
We used this approach to validate our models and comparing these models with the observed movements instead of the standard Viterbi predictions by the Viterbi algorithm because Viterbi predictions, which use the most probable sequence of movement states based on the observations \cite{zucchini_hidden_2009,langrock_flexible_2012}, double-count the observed data. It is useful to predict missing data in the observation sequence, but because it is conditional on the observed values, it can not reliably evaluate goodness of fit for the different structural complexities of HMM models.

@@

Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is somewhat overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

The panthers exhibits a clear diurnal pattern from the average hourly step lengths from the observed data (Figure 4). As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle. All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns. However, the block model does markedly worse than the other temporal models (changing the block definitions might help), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough). We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points).

@@

We would like to thank Madelon van de Kerk, Madan Oli, and David Onorato for their previous work on Florida panthers. We also would like to thank McMaster University, Florida Fish and Wildlife Conservation Commission and many individuals for data collection and fieldwork. Lastly, to Madelon van de Kerk for making the data available at the Institutional Repository at the University of Florida (IR@UF).

@@

\section*{Tables}
\begin{table}[h!]
\caption{Cat ID}
      \begin{tabular}{cccc}
        \hline
           & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
         & 130 & 1 & 10286\\
         & 131 & 2  & 9458\\
         & 48 & 14  & 14645\\
         & 94  & 15   & 10250\\ \hline
      \end{tabular}
\end{table}

>>>>

\startlocaldefs
\newcommand{\obs}{\mathbf{y}}
\newcommand{\cov}{\mathbf{x}}
\newcommand{\state}{z}
\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}
\newcommand{\bmb}[1]{\textbf{[BMB: #1]}}
\endlocaldefs

@@

\parttitle{Background} 
Clustering time-series data into discrete groups can improve 
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in 
probabilities of group occupancy, or the rates at which individuals move between groups, can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model long-term hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal heterogeneity in transition probabilities, a straightforward 
but little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers.

\parttitle{Results}

Simulations and analyses of panther data showed that model
misspecification (omitting important sources of variation) can lead to
overfitting/overestimating the underlying number of behavioural
states.  Models incorporating temporal heterogeneity identify fewer
underlying states, and can make out-of-sample predictions that
capture observed diurnal and autocorrelated movement patterns exhibited by
Florida panthers.

\parttitle{Conclusion}

Incorporating temporal heterogeneity improved goodness of fit and predictive
capability as well as
reducing the selected number of behavioural states 
to a more biologically interpretable level. Our results
suggest that incorporating additional structure in statistical models 
of movement behaviour can
allow more accurate assessment of appropriate model complexity.

@@

Given a sequence of animal movements, movement models aim to find a parsimonious description
that can be used to understand past movements and predict
future movements. Ecologists have long considered the effects of
individual-level covariates (sex, age, nutritional status) and
environmental covariates (habitat type, location of predators or prey)
on movement \cite{patterson2008state, mckenzie2009first, pal1998dispersal}. 
More recently, modelers have developed
\emph{hidden Markov models} (HMMs)
\cite{firle_influence_1998,nathan_movement_2008,langrock_flexible_2012} % 
--- used in animal ecology under the rubric of the ``multiphasic movement
framework'' \cite{fryxell_multiple_2008} --- that consider the effects
of organisms' \emph{internal} states; in particular, HMMs model animal
movement as though individual animals' movement behaviour 
at particular times is determined
by which of a discrete set of unobserved movement states
(e.g. ``foraging'', ``traveling'', ``resting'') they currently occupy.
Conditional on the state occupied by an individual, HMMs typically
assume that animals follow a standard correlated random walk model
\cite{okubo_diffusion_1980,turchin1998quantitative}.

@@

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in transition
probabilities based on
the local environment can be accounted for by incorporating environmental
covariates in the HMM \cite{patterson_classifying_2009}, or 
inferred from direct comparisons between inferred states and environmental
conditions \cite{fryxell_multiple_2008}. Schliehe-Diecks et al. 
\cite{schliehe-diecks_application_2012}
considered temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales \cite{gurarie2009novel}.

@@

van de Kerk et al. \cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, Dean et al.~\cite{dean_behavioural_2013} evaluated models with up to 10
states, but like van de Kerk et al. they chose to consider only 
models with three states. 
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

Our second goal, therefore, is to explore whether
van de Kerk et al.'s results on optimal
model complexity might be driven at
least in part by structural problems with their statistical
model, i.e. the
assumption of temporally homogeneous behaviour.  For large data sets, 
information-theoretic model selection methods will
typically choose complex, highly parameterized models; when there is
only one way in which models can become more complex (e.g. by
increasing the number of latent states), complexity that is
present in the data but not accounted for in the model (e.g. spatial
or temporal heterogeneity) can be misidentified as other forms of
complexity.  We predict that increasing volumes of data will
increasingly lead researchers who are accustomed to fitting 
small models to sparse data into such traps.  We examine whether
allowing for diurnal variation in the Florida panther data allows us to
select models with fewer latent states; we also
fit models to simulated data with varying numbers of latent states,
and with and without temporal heterogeneity, to test our conjecture that
heterogeneity can be misidentified as behavioural complexity.

@@

@@

\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \times \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\begin{description}
\item[Initial probability] $P(\state_{1} = i | \cov_{1}) P(\obs_{1} | \state_{1}, \cov_{1} )$: the probability of state $i$ at time $t=1$ given that the covariates are $\cov_{1}$, times the vector of observations $\obs_{1}$ conditioned on state $\state_{1}$ and covariates $\cov_{1}$.

\item[Transition probability] $P(\state_{k}=j | \state_{k-1}=i,\cov_{k})$: the probability of a transition from state $i$ at time $t=k-1$ to state $j$ at time $t=k$, given covariates $\cov_{k}$ .

\item[Emission probability] $P(\obs_{k} | \state_{k},\cov_{k})$: a vector of observations $\obs_{k}$ given state $\state_{k}$ at time $t=k$ and covariates $\cov_{k}$ .
\end{description}

@@

For an $n$-state HMM, we need to define an $n \times n$ matrix that specifies the probabilities $\pi_{ij}$ of being in movement states $j$ at time $t+1$ given that the individual is in state $i$.  The FMM is a special case of HMM where the probabilities of \emph{entering} a given state are identical across all states --- i.e., the probability of occupying a state at the next time step is independent of the current state occupancy. It can be modelled in the HMM framework by setting the transition probabilities  $\pi_{ij} = \pi_{i*}$.

@@

where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block corresponding to time $t$, and 0 otherwise).

\item[Quadratic transition model] 
We assume the elements of the linear predictor
are quadratic functions of hour:

$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2} .
$$

The quadratic model is not diurnally continuous, i.e. there 
is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity 
constraint would collapse the model to a constant.

\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint:

$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right) .
$$

\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day. This model is included for comparative purposes; due to the large number of parameters in the model (more than $24n(n-1)$ for a HMM with $n$ states), it is not really practical. We only fitted up to four states using the hourly model.  
\end{description}

Other periodic functions, such as Fourier series (i.e., the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could also be considered.

@@

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
within each type of transition complexity \bmb{clarify?} ($\dbic$ = BIC - min(BIC)) assesses the overall 
goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
observed data and for data simulated from the models shows how well a particular class of 
models can capture diurnal variation in behaviour. (3) Comparing temporal autocorrelations
for the observed data and for data simulated from the models shows how well a particular class
of models can capture serial correlation at both short and long time scales.

Model complexity and the number of parameters increase as the number of latent states increase.
For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity. In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:

\begin{equation}
\textrm{Number of Free Parameters}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}} .
\label{eq:numparam}
\end{equation}

As the number of states increases, the number of free parameters in (homogeneous or heterogenous) FMMs and time-homogeneous HMMs will increase linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically (Eq.~\ref{eq:numparam}). When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters. \bmb{what does last sentence mean?? delete?}

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
because the states predicted by the Viterbi algorithm already rely on the observed data. This approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for the different structural complexities of HMM
models.
\bmb{we might need to be more careful here. Apparently Zucchini/Langrock et al also define ``pseudo-residuals'';
can you look these up and see if that approach suffers from the same issues as Viterbi ... ?}

@@

Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

The average hourly step lengths from the observed panther data exhibit a clear diurnal pattern (Figure 4). As expected, temporally homogeneous models (whether FMM or HMM) predict the same mean step length regardless of time of day, failing to capture the diurnal activity cycle. All of the models incorporating temporal heterogeneity, including the temporally heterogeneous FMM, can capture the observed patterns. However, the block model does markedly worse than the other temporal models (changing the block definitions might help \bmb{clarify?}), and the (overparameterized) hourly model does better than any other model at capturing the early-evening peak (but worse at capturing the mid-day trough). We also included average hourly step lengths from three-state temporally homogeneous HMM Viterbi prediction (v points \bmb{clarify?}).

@@

We would like to thank Madelon van de Kerk, Madan Oli, and David Onorato for their previous work on Florida panthers. We also would like to thank McMaster University, Florida Fish and Wildlife Conservation Commission and many individuals for data collection and fieldwork. Lastly, we thank Madelon van de Kerk for making the data available at the Institutional Repository at the University of Florida (IR@UF).

@@

\section*{Tables}
\begin{table}[h!]
\caption{Cat ID and number of observations; ID numbers are given matching
those shown by van de Kerk et al. 2014 and those 
in the data located at the UF Institutional repository (IR@UF).}
      \begin{tabular}{cccc}
        \hline
           & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
         & 130 & 1 & 10286\\
         & 131 & 2  & 9458\\
         & 48 & 14  & 14645\\
         & 94  & 15   & 10250\\ \hline
      \end{tabular}
\end{table}
